Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-01-28 paper/main.tex
Updates
updates

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remains limited, as researchers often prefer to construct DAGs manually
based on their expert knowledge. This preference arises due to practical issues
with these automated algorithms, such as making obvious errors and outputting
Markov Equivalence Classes (MECs). To address these challenge, we propose an
iterative method to assist researchers in manually constructing or modifying
DAGs by combining implied Conditional Independence (CI) testing with expert
knowledge. We introduce a novel measure of partial association for mixed data
to rank violations of implied CIs, enabling users to prioritize model
modifications while leveraging their expertise to decide the orientation of
edges when making these modifications. Empirical results show that this
approach achieves performance comparable to automated algorithms, assuming the
user can correctly orient edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model can
potentially determine edge orientations.

@@

Figure~\ref{fig:intro} shows an example to highlight some of these issues using
the Adult income dataset \citep{Becker1996}. As can be seen in the figure
depending on the sample size and the algorithm used, we get vastly different
results.

The other ways is to do post hoc modification to the learned model using
implied CI based model testing \citep{Ankan2021, Ankan2023}. In this testing
procedure, users test whether the implied CIs of the DAG hold in the dataset or
not using statistical CI tests. Based on the results of these statistical
tests, we can accordingly make modifications to the current model to fix the
violations. However, this CI based testing also has a few drawbacks: 1) CIs are
implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

@@

% using a conditional measure of
% association.  To deal with both this issues, in this paper, we develop a
% measure of partial association which can be using in combination of the CI test
% to rank the violations in these CI tests. On the other hand, in Structural
% Equation Models (SEMs) literature, more emphasis has been placed on developing
% tools to help researchers manually construct models while integrating their
% domain expertise. Our proposed approach uses the measure of association to find
% correlated variables in the data that are not explained by the current model.
% Using the measure, we can then rank the pair of variables between which adding
% an edge would improve the explanation the most. Based on this information, the
% user can select a potential pair of variables and based on their expert
% knowledge decide the direction of the edge.

\begin{figure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=1]{figures.pdf}
\end{subfigure}%
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=2]{figures.pdf}
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=3]{figures.pdf}
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=4]{figures.pdf}
\end{subfigure}

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remains limited, as researchers often prefer to construct DAGs manually
based on their expert knowledge. This preference arises due to several
practical issues with these automated algorithms, such as making obvious errors
and outputting Markov Equivalence Classes (MECs). To address these challenges,
we propose an iterative method to assist researchers in manually constructing
or modifying DAGs by combining implied Conditional Independence (CI) testing
with expert knowledge. We introduce a novel measure of partial association for
mixed data to rank violations of implied CIs, enabling users to prioritize
model modifications while leveraging their expertise to decide the orientation
of edges when making these modifications. Empirical results show that this
approach achieves performance comparable to automated algorithms, assuming the
user can correctly orient edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model can
potentially determine edge orientations.

@@

Fig.~\ref{fig:intro} highlights some of these issues using the Adult income
dataset \citep{Becker1996}. As can be seen in the figure depending on the
sample size and the algorithm used, we get vastly different results.

As a result of these challenges users have relied on model testing approaches
along with their expert knowledge to construct these models. The other ways is
to do post hoc modification to the learned model using implied CI based model
testing \citep{Ankan2021, Ankan2023}. In this testing procedure, users test
whether the implied CIs of the DAG hold in the dataset or not using statistical
CI tests. Based on the results of these statistical tests, we can accordingly
make modifications to the current model to fix the violations. However, this CI
based testing also has a few drawbacks: 1) CIs are implied only by missing
edges in the model, so we can not test for already existing edges. 2) In high
sample size scenarios, p-values can suggest a significant association between
variables even if they have a very weak relationship. 3) Number of implied CIs
can be large making it difficult to manually go through each of them; for
example, in the moderately sized alarm network with $ 37 $ node, there are
$287$ implied CIs.

@@

\begin{figure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=1]{figures.pdf}
	\caption{PC algorithm $ \chi^2 $ test. $ N=400$ }
\end{subfigure}%
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=2]{figures.pdf}
	\caption{PC algorithm $ \chi^2 $ test. $ N=800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=3]{figures.pdf}
	\caption{PC algorithm residualization test. $ N=400$ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=4]{figures.pdf}
	\caption{PC algorithm residualization test. $ N=800$ }
\end{subfigure}
\begin{subfigure}{0.25 \textwidth}
	\includegraphics[page=5]{figures.pdf}
	\caption{Hill Climb Search BIC Score. $ N=400$ }
\end{subfigure}%
\begin{subfigure}{0.25\textwidth}
	\includegraphics[page=6]{figures.pdf}
	\caption{Hill Climb Search BIC Score. $ N=800$ }
\end{subfigure}
